---
title: "R Programming Tips"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    toc: yes
    css: "menu.css"
---
```{=html}
<html lang="en">
<head>
  <script src="https://cdnjs.cloudflare.com/ajax/libs/jquery/3.6.0/jquery.min.js"></script>
  <link rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.15.3/css/all.min.css"/>
  <link rel="stylesheet" href="menu.css">
</head>
<body>
  <header>
    <div class="inner-width">
      <div class="menu-icon">
        <i class="fas fa-align-right"></i>
      </div>
    </div>
  </header>

  <div class="navigation-menu">
    <nav>
      <li><a href="index.html">Home</a></li>
      <li><a href="research.html">Research</a></li>
      <li><a href="resources.html">Resources</a></li>
      <li><a href="CV.html">CV</a></li>
      <li><a href="contact.html">Contact</a></li>
    </nav>
  </div>

  <script>
    $(".menu-icon").click(function(){
      $(this).toggleClass("active");
      $(".navigation-menu").toggleClass("active");
      $(".menu-icon i").toggleClass("fa-times");
    });
  </script>
</body>

</html>


```


```{r include=FALSE}
library(tidyverse)

```



## Preamble

Over the past couple of years, I've had the privilege to advance my R skills, as well as acquire useful functions that should aid anyone using R for behavioral science. This list is not exhaustive, but a list of my most used functions, packages, and useful tips!

## Pipe `%>%` Operator

The pipe operator, written as `%>%` takes the output of one function and passes it into another function as an argument. This allows us to link a sequence of analysis steps. 

\n For a mathematical analogy, `f(x)` can be rewritten as `x %>% f`

```{r echo=TRUE}
## compute the logarithm of `x`

x <- 1

log(x)

## compute the logaritm of `x`

x %>% log()


```

Why is this useful though? R is a functional language, which means that your code often contains a lot of parenthesis, `(` and `)`. When you have complex code, this often will mean that you will have to nest those parentheses together. This makes your R code hard to read and understand.

```{r echo=TRUE}
# Initialize `x`
x <- c(0.109, 0.359, 0.63, 0.996, 0.515, 0.142, 0.017, 0.829, 0.907)

# Compute the logarithm of `x`, return suitably lagged and iterated differences, 
# compute the exponential function and round the result
round(exp(diff(log(x))), 1)

# Compute the same computation as above but with pipe in operator

x %>% log() %>%
    diff() %>%
    exp() %>%
    round(1)

```

In short, here are four reasons why you should be using pipes in R:

1. You'll structure the sequence of your data operations from left to right, as apposed to from inside and out;

2. You'll avoid nested function calls;

3. You'll minimize the need for local variables and function definitions;

4. You'll make it easy to add steps anywhere in the sequence of operations.

## dplyr package
 
By far my most used package in R is `dplyr`. See documentation [here](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8) 


